How is experience transformed into memory?

### 2018-08-28 paper/main.tex
discussion edits

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape. In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g.\ percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

Prior work on neural responses during naturalistic experiences and recall has largely focused on identifying brain regions whose responses are reliably similar across individuals ~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17}. This allows one to identify which regions might be processing or representing the stimulus, or retrieving details of the experience during recall. Our approach is fundamentally different. We ask: which brain regions during movie viewing match the movie structure and how individual participants will recall the movie later (in terms of the temporal correlations of their recall topic trajectory)? This latter approach highlights regions such as the medial prefrontal cortex (mPFC) and medial temporal lobes (MTL) that may respond in idiosyncratic ways across individuals, but that nonetheless play an important role in encoding experience into memory. In other work, the mPFC has been suggested to play a role in the representation of a ``schema'', or a network prior knowledge that is consistent with the current task ~\citep{KestEtal12, GilbMarl17}. One interpretation for mPFC's involvement in our work is that mPFC represents participant-specific relationships with the content of the video, and its representation during scene viewing is related to how a participant later describes the scene. MTL activity patterns also matched participant's recall, which is consistent with a large body of literature highlighting the role of the MTL in episodic memory encoding ~\citep{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, such as the classroom or in a doctor's office.  Whereas tests often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might miss the ``forrest for the trees''. We view ``true learning'' as understanding the key concepts (i.e., understanding key themes in the learned content and how they relate) rather than about regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of a memory disorder, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the movie's topic trajectory? Or if participants were told that their narrations would be played back to other participants ~\citep{ZadbEtal17}, would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events or with more elapsed time between encoding and recall, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing out" of the trajectories with each new retelling.

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g.\ one that describes the video in the same way as participant's recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g.\ participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to details (while maintaining flexibility).

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While its undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for naturalistic experiences. Real-life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory should not ignore this fact. Our work provides a theoretical advance in our understanding of what it means to remember as well as a novel methodological tool to study it.

@@

\section*{Additional measures of naturalistic memory}
To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered and computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p<.05$, participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r(16)=-.57, p=.016$). Thus, a closer match between the video and recall event models was related to better recall performance.

Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event indexing how similar the recall event was to its matching movie event (i.e the ``precision'' of the recall). We then averaged the correlations within participant. In line with our prediction, there was a strong correlation between hand annotated memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's $r(16)=.74, p=.0006$).

@@

In order to ascertain which features were important in relating the recall models to the video model, we similarly compared to the temporal structures of video and recall models deprived of a single feature at a time. For each feature removed, we transformed each subject’s recall transcript using a model trained on the feature-deprived video text windows (of 50 scene segments), and resampled the recall timeseries to match the shape of the movie model (1976 timepoints). We then represented the temporal structures of each subject’s recall model as timepoints (1976) by timepoints (1976) correlation matrices and computed the average correlation with the temporal structure of the video model (across subjects).

>>>>

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes recollection of an experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the video participants viewed by measuring the match between the shapes of the video's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory -- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape. In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g.\ percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

Prior work on neural responses during naturalistic experiences and recall has largely focused on identifying brain regions whose responses are reliably similar across individuals ~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17}. This allows one to identify which regions might be processing or representing the stimulus, or retrieving details of the experience during recall. Our approach is fundamentally different. We ask: which brain regions during video viewing match the video structure and how individual participants will recall the video later (in terms of the temporal correlations of their recall topic trajectory)? This latter approach highlights regions such as the medial prefrontal cortex (mPFC) and medial temporal lobes (MTL) that may respond in idiosyncratic ways across individuals, but that nonetheless play an important role in encoding experience into memory. In other work, the mPFC has been suggested to play a role in the representation of a ``schema'', or a network of prior knowledge that is consistent with the current task ~\citep{KestEtal12, GilbMarl17}. One interpretation for mPFC's involvement in our work is that mPFC represents participant-specific relationships with the content of the video, and its representation during scene viewing is related to how a participant later describes the scene. MTL activity patterns also matched participant's recall, which is consistent with a large body of literature highlighting the role of the MTL in episodic memory encoding ~\citep{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, such as the classroom or in a doctor's office.  Whereas academic tests often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might ``miss the forrest for the trees''. We view ``true learning'' as understanding key concepts (i.e., understanding central themes in the learned content and how they relate) rather than regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of memory disorders, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, and other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the video's topic trajectory? Or if participants were told that their narrations would be played back to other participants ~\citep{ZadbEtal17}, would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events or with more elapsed time between encoding and recall, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing out" of the trajectories with each new retelling.

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g.\ one that describes the video in the same way as participants recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g.\ participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to such details while maintaining flexibility.

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While it's undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for real-life experiences. Real-life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory must not ignore this fact. Our work provides a theoretical advance in our understanding of what it means to remember as well as a novel methodological tool to study it.

@@

\section*{Additional measures of naturalistic memory}
To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the video model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered and computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p<.05$, participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r(16)=-.57, p=.016$). Thus, a closer match between the video and recall event models was related to better recall performance.

Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event indexing how similar the recall event was to its matching video event (i.e the ``precision'' of the recall). We then averaged the correlations within participant. In line with our prediction, there was a strong correlation between hand annotated memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's $r(16)=.74, p=.0006$).

@@

In order to ascertain which features were important in relating the recall models to the video model, we similarly compared to the temporal structures of video and recall models deprived of a single feature at a time. For each feature removed, we transformed each subject’s recall transcript using a model trained on the feature-deprived video text windows (of 50 scene segments), and resampled the recall timeseries to match the shape of the video model (1976 timepoints). We then represented the temporal structures of each subject’s recall model as timepoints (1976) by timepoints (1976) correlation matrices and computed the average correlation with the temporal structure of the video model (across subjects).
